Ribosome accumulation during early phase resistance training in humans

### 2022-02-04 response-to-reviews-rev1.Rmd

---
title: "APH-2021-12-0652: Responses to reviewers comments"
author: 
  - Daniel Hammarström
  - Sjur J. Øfsteng
  - Nicolai B. Jacobsen
  - Krister B. Flobergseter
  - Bent R. Rønnestad
  - Stian Ellefsen
date: "2022-02-04"
output: html_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliogPlease find our raphy: references.bib
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Response to decision letter

**Authors answer**: We would first like to thank both reviewers for their interest in our manuscript and their questions and constructive criticism. We feel that the revised manuscript has benefited from this first round of reviews. We have addressed the reviewer's comments below. We have updated the figures to match the journals guidelines, we are sorry not to have noticed the very clear instruction prior to the first submission. We have spotted some grammatical errors and changed these. We also noted that information was missing regarding one missing biopsy from one participant at baseline and technical difficulties in the calibrator sample of one participant for protein analyses. The latter affecting the sample size in a subset of analyses (n = 10 instead of 11; see revised manuscript Methods and Materials and Table 1).

## Reviewer 1

> **Reviewer comment:** This manuscript investigates the induction of ribosome biogenesis and
> accumulation of RNA/ribosomes during the early phase of resistance
> training. Subjects performed unilateral RE bilaterally but with
> differing loading protocols. One leg trained with a constant volume
> whereas the other trained with variable volume. Control subjects were
> biopsied at relevant time points. The manuscript presents evidence in
> favor of that skeletal muscle in the naïve state is very responsive
> with respect to ribosome biogenesis following acute RE. With repeated
> training sessions this induction is maintained but not further
> stimulated. Most of the effect on total RNA was seen during the first
> training block. UBF levels were related to RNA accumulation and rate
> of total RNA accumulation predicted RT-induced muscle hypertrophy.
> With eight days of detraining, around 20% of total RNA was lost,
> indicating that the elevated RNA pool needs to be maintained not to
> decrease. The manuscript is highly interesting and ambitious in its
> design. I think that it adds to the published literature and could
> potentially be a future important reference piece. However, I have
> some concerns and comments that I'd like for the authors to address,
> as I think that they can improve the manuscript further.

> The western blot membranes in the paper does not really translate into
> the presented data. In my opinion the bands look rather faint and
> equal in intensity across the time course. Could the authors perhaps
> provide raw data and membranes for rpS6 and UBF just for the
> reviewers?

@@

  group_by(target) %>%
  summarise(cv = mean(cv, na.rm = TRUE)) %>%
  pull(cv)

cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")

```

**Authors answer:** We agree with the referee that the blots presented in Figure 2 were faint, especially for UBF. The selected blot was made for illustrative purposes as we had ordered the samples according to sampling time and condition in that particular blot. Unfortunately this blot was in the "fainter end" of the spectrum. We have replaced the blot with one that we feel is more representative. We are providing the quantified blots in a separate document for the reviewers.

The raw aggregated data (averaged over replicates per samples) does show a clear increase in signal over the time course (Figure 1 in supporting document). We acknowledge that the western blot technique is associated with considerable technical variation. The average coefficient of variation (CV) in replicates were `r cvs` for rpS6 and UBF respectively, both within the
expected range of CV from western blot data [@degasperi2014].}

@@

  theme(panel.background = element_rect(fill = "#fefeda"), 
        panel.grid.major.y = element_line(color = "gray80"), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

plot_grid(absolute_data, relative_data, 
          ncol = 1, rel_heights = c(0.55, 0.45))

```

> **Reviewer comment:** Is pre-rRNA 47S ETS and 45S ETS per unit tissue weight a previously reported way to present qPCR data? Does the data look the same in relation to 1 or more house keeping genes?

**Authors answer:** One transcribed pre-rRNA results in a single end product, one ribosome, and in the context of protein synthesis, the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis [@millward1973; @west2016]. In contrast to rRNA processing, mRNA transcription and subsequent translation could, to some degree, be seen as competitive between different mRNA. This translates to different normalization strategies where we believe that the most biologically relevant perspective on rRNA is expressing it to the amount of tissue probed in the qPCR. The use of a housekeeping or reference gene in this respect could be misleading as the relative stability of the reference gene must be determined to some quantity of interest. Normalization of mRNA data using reference genes assumes that such stability has been determined as relative to the mRNA pool [@andersen2004], or any other quantity. We have however noted that in response to RT, relative stability is hard to determine as the whole mRNA pool is increasing [@khan2020]. This again brings us to the basic problem of normalization, what is the underlying quantity of interest? We have not attempted to determine relative stability of reference genes in the present data set because of our past experiences with the RT model, therefore we are unwilling to explore the effect of using a reference gene, as the reviewer suggests. A simple alternative to determining the relative stability of a reference gene is to use an external (exogenous) reference together with the known quantity of tissue used in sample preparations. The use of exogenous reference genes has previously been used to achieve "custom normalization" to increase biological relevance (e.g. [@ellefsen2008]). We have also previously used the same approach as in the present work to conclude that RT leads to increased abundance of pre-rRNA and rRNA in muscle tissue [@hammarström2020].

> **Reviewer comment:** Could the authors please comment on the apparent disconnect between increase in RNA and rpS6 protein levels?

**Authors answer:** We thank the reviewer for raising this question as we have attempted to broaden our discussion. We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and rpS6. The model used "controls" for both quantities being increased over the whole training period, within a specific time-point there exists no relationship between rpS6 and total RNA. This separates our analysis from others who have shown correlations between rpS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase. As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and rpS6 protein levels. Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam2007], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis. We have expanded the discussion by raising the point that unbound ribosomal proteins (including rpS6) may be passively increased by myonuclear accretion. Although somewhat speculative, this notion is supported by recent evidence suggesting a specific role of satellite cell fusion in increasing availability of ribosomal proteins together with increased rRNA transcription [@murach2022]. This could in part help explain our observations since an increased myonuclear pool is presumably maintained during short-term de-training but rRNA transcription is decreased (and/or ribophagy increased).

@@

Our data are however limited to a single ribosomal protein without data supporting accretion of myonuclei. This limitation was also added to the discussion.

> **Reviewer comment:** The concept that the increase in RNA content predicts subsequent hypertrophy is very interesting. Even more intriguing is the finding that total RNA mid training negatively correlates to muscle hypertrophy is fascinating! I would encourage the authors to deepen the discussion on these key data points, especially in the light of more and more evidence in favor of ribosome specialization. Could a high percentage of new ribosomes be advantageous for growth in more ways than just contributing greater numbers?

**Authors answer:** The reviewer raises an interesting point, that we interpret as suggesting that newly synthesized ribosomes would be more beneficial for muscle growth in response to a new stimuli as novel ribosomes are potentially specialized for growth rather than cellular status quo. In light of recent studies and overviews highlighting ribosomal heterogoneity, our results are worth following up in relation to a specialized ribosome.

@@

We have added a section in the discussion highlighting this aspect of the results.

> Please comment on that UBF levels where maintained at the detraining time point but total RNA decreasing. Would it be relevant to analyze UBF phosphorylation?

**Authors answer:** Although not robustly decreased, the overall tendency of total UBF protein was decrease in response to the de-training period. The variation presumably contributed to relationship with total RNA abundance. However, as the reviewer points to, we did not measure the "active" UBF nor UBF in the context of rDNA chromatin regulation. It is possible that in the de-training period a reduced stimuli (possibly through reduced mTORC1 stimulation [@nader2005]) reduced interaction between UBF and rDNA. Unfortunately, a limited amount of tissue (together with other resources) prevented us from doing more elaborate analyses, including UBF phosphorylation or UBF/rDNA interactions. We noted this limitation in the discussion.

> **Reviewer comment:** The fact that RNA decreased at the detraining time point is interesting and indicates the importance of actively maintaing translational capacity not to loose it. I think the discussion could benefit from including a short section of the role for ribophagy in this setting?

**Authors answer:** We thank the reviewer for this suggestion, we have incorporated a section connecting our discussion to the emerging field of ribophagy.

> **Reviewer comment:** The methods in the abstract must be improved. In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I'd like for the authors to try and specify biopsy time points.

**Authors answer:** We thank the reviewer for catching this typo. We have updated the abstract with the correct information and additional information regarding biopsy time points.

> **Reviewer comment:** Page 3, line 48-49. I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys). This paper shows the specific response of ribosome biogenesis in response to RE and not Endurance type exercise.

**Authors answer:** We thank the reviewer for this suggestion.

**Reviewer comment:**
> Page 4, line 8: Please clarify "numberical lowering". Could this be rephrased?
> Page 4, line 10: Double ".."
> Page 4, line 15: Double ".."
> Page 4, line 26: Should "to" be removed?
> Page 11, line 20: n=7 or 8? Please check abstract
> Page 11, line 35: Should "a" be removed? Please double check if it's 7 or 8 days of detraining. Abstract states 8.

**Authors answer:** We thank the reviewer for pointing out the above typos, they have been corrected.

**Reviewer comment:**
> Page 12, line 19: please clarify "leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?

**Authors answer:** Yes, sets were performed with rest periods for the other leg. The protocol still allowed for complete rest between sets ($\sim$ 60 sec) and adjustment of loads if necessary. A clarification has been added to this section.

**Reviewer comment:**
> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen. What happened to the other material? If I misunderstood, can the text be rewritten for clarity?

**Authors answer:** Muscle tissue from all passes were processed together and aliquots were weighed and frozen. The section has been updated. Unfortunately, a malfunctioning freeze-dryer destroyed a portion of the material. The second aliquot was pooled for RNA extraction when available.

**Reviewer comment:**
> Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.

**Authors answer:** The spike-in was used to for normalization of rRNA to muscle weight as previously described [hammarström2020]. The section has been updated to notify the reader of the purpose of adding the external reference.

**Reviewer comment:**
> Page 14, line 49: Does "normalized" refer to denatured?

**Authors answer:** This refers to diluting all samples to a common protein concentration. The sentence has been updated for clarity.

**Reviewer comment:**
> Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?

**Authors answer:** Yes, the sentence should read amplicon sizes. This has been updated.

**Reviewer comment:**
> Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?

**Authors answer:** The figure legends has been updated together with an overall update of figure appearances.

## Reviewer 2

**Reviewer comment:**
> The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training. Phenotype variables were also assessed. Training led to increases in muscle growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6. Training volume did not play a role in these adaptations. Interestingly, the rate of total RNA increases predicted hypertrophy. Likewise, training cessation led to decrements in total RNA abundance. I love this study. This continues to contribute to our knowledge regarding the role of ribosome biogenesis in muscle hypertrophy, and is delicate human work with multiple time point biopsies. I have minor comments in order of appearance, and congratulate the authors on their fine work.

**Authors answer:** We are grateful for the reviewers kind words.

**Reviewer comment:**
> ABSTRACT: Reads well and concise. Nice work, and no comments.

@@

> Page 4 - Lines 10 and
> 15, minor: each sentence has two stops.

> Line 33, minor: no need to use the phrase "per se"

> RESULTS - Line 51 (and throughout the results), clarification: is "9.2%-point difference" the same as "9.2% difference"? If so, adjust and remove "point" from all descriptors as
this is confusing.

**Authors answer:** We have corrected in line with the reviewers notes. The "%-point difference" refers to a differences in percentage points, i.e. one condition increases in 2% the other 11.2%, the percentage-point difference is 9.2%. This is different from where percentages are stated without the "-point". Although confusing we believe that it is the correct way of reporting this such results.

**Reviewer comment:**
> Other than this, the results, while dense and full of data, are well-written.

@@

> DISCUSSION Page 7 - Line 17, minor: again, "per se" can be removed. - 
> Line 19, minor: "was not affected" better reads as "were not affected" 
> Line 39, minor: "session" should read as "sessions"

**Authors answer:** We thank the reviewer for spotting mistakes and suggesting improvements to the text. These have been corrected.

**Reviewer comment:**
> I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56. We've done a lot of work in the ribosome biogenesis area, and we've thought a good bit about total RNA/unit muscle with regard to what it represents. The authors correctly articulated this as a proxy of ribosome density. However, as muscle tissue and myofibers volumetrically increase, this will affect the metric (the same holds true with mitochondrial biogenesis and using CS activity as a proxy of mitochondrial volume density). An easy way to interpret this is in the following example: o Muscle tissue increases 5% in thickness with RT (and myofibers scale accordingly as well) o Total RNA/unit muscle does not change in value Most would interpret this as ribosome biogenesis not occurring. However, biogenesis likely occurred because, as muscle grew, ribosome biogenesis scaled with growth thus leading to no change in RNA concentration. Our laboratory discusses this important concept with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID: 34646153).

@@

> Anyhow, this is no critique of the authors' hard and excellent work. However, I think the authors re-iterating this point as a stand-alone paragraph would benefit readership. In addition, I believe taking a moment to define what the total RNA/ribosome density metric conceptually represents, and then (in this newly added paragraph) briefly walking the reader through each phase of the data, and what could be going on with biogenesis in the context of a growing muscle would be interesting to see. For instance, in Figure 3G, my interpretation (much like the authors') is that ribosome biogenesis is very rapid at the onset of RT where little functional growth is likely occurring. And like the authors, I believe this likely was needed for eventual growth to occur. However, the "no change" in RNA concentrations session 9-12 as muscle is continuing to grow may not indicate a plateau in biogenesis given that this is the time during which muscle growth is likely starting to ramp up. Hence, biogenesis rates could be equally as ramped here compared to sessions 1-9 of RT.

**Authors answer:** We thank the reviewer for this suggestion. We agree with the reviewer and feel that the manuscript also communicates this message. This is an important point to underline. We felt that the section served well as part of the mentioned section and added the following:

\textcolor{red}{"This may indicate that during the plateau phase, the ribosomal concentration is balanced by muscle growth. This balance, measured as a constant ribosomal density in a growing cell, still requires the biosynthesis of ribosomes to match the volumetric expansion of the cell. As such, indirect measures of translational capacity such as the concentration of total RNA may mask the absolute increase in ribosomes that occurs during periods of muscle hypertrophy."}

**Reviewer comment:**
> METHODS
> Page 11
> -   Lines 9-19: Authors stated that 18 participants were recruited, but
    then go on to say that n=11 TRAIN and n=8 CTRL completed. Please
    reconcile.
> -   Line 35: Authors stated 7 days after detraining, but in the
    discussion (page 7, line 19) they mention after 8 days of
    detraining. Please reconcile. Page 12
>   Line 39, minor: "restitution" should read as "rest". Please change
    accordingly, and if used throughout the paper, please alter. Page 15
>   Line 36, minor: "Glycin" should read as "Glycine". In addition, and
    molar unit is needed.

**Authors answer:** The above points has been addressed in the revised manuscript, thank you for your detailed reading of the manuscript.

### 2022-02-04 manuscript.Rmd
revision 1 complete

<div custom-style="runningtitle_style">**Running title:** Ribosome biogenesis in early phase resistance training</div>

@@

**Aim:** To describe the time course of ribosome accumulation during resistance training, its relation to training volume, UBF protein levels, and muscle growth.

**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. <span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9 and 12</span> and after eight days of de-training, as well as</span> from a non-training control group (n = 8). In addition, muscle thickness was measured before and after training and control periods.

**Results:** Training led to muscle growth (`r hypertrophy` vs. control) with concomitant increases in total RNA, rRNA, UBF, and RPS6 with no differences between volume conditions. Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline). After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session). After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.

@@

**Keywords:** muscle hypertrophy, resistance training, ribosome biogenesis

@@

Skeletal muscle is a critical target for interventions that promote health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
with resistance training (RT) being the advocated remedy.  
Prolonged RT leads to changes in the balance between muscle protein breakdown and synthesis, with one bout of resistance exercise acutely increasing protein synthesis for up to 48 hrs after exercise [@RN786],<!-- Phillips 1997 -->
<span custom-style="CHANGE">and</span> subsequent repeated bouts leading to accumulation of muscle protein over time
[@RN1272<!-- Phillips review 2014 -->;
@RN1642]<!-- Brook 2015 experiment -->. 
In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased <span custom-style="CHANGE">basal muscle protein synthesis rates</span>
[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
@RN1866;<!-- Wilkingson 2008 -->
@RN1897],<!-- Reidy 2017 Increased rested state protein synthesis -->
which has been postulated to be associated with increased translational capacity, i.e. accumulation of ribosomes
[@RN1897;<!-- Reidy 2017 -->
@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, which is closely connected to protein synthesis
[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
and muscle hypertrophy
[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
@RN1755; <!-- Stec Ribosome bio higher in high responders -->
@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1810]<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.

Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
[@RN1920;<!-- Warner review yeast -->
@RN1820;<!-- Moss review -->
@RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
[@RN1820]<!-- Moss review -->.
Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
Such activation is at least partly controlled by the mechanosensitive mTOR pathway, with its inhibition being associated with blocked UBF phosphorylation and subsequent rRNA transcription
[@RN1632; <!-- Nader rapamycin block UBF availability  -->
@RN2564].<!-- Hannan mTOR  -->
Interestingly the availability of UBF *per se* has been shown to be a determinant of rRNA transcription
[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->

Resistance exercise is a potent <span custom-style="CHANGE">and specific [@figueiredo2021]</span> stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
[@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->].
Repeated bouts lead to the accumulation of mature rRNA reflected in total RNA and presumably functional ribosomes
[@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
@RN1897<!-- 11% 36 sessions young -->].
However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with a mere few studies having investigated exercise-induced changes in rRNA over multiple time-points, all of which are either limited to a selected few time-points or a limited time frame.
For example, two consecutive bouts of electrically evoked muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
[@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
Using voluntary contractions, peak values were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809],
resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), <span custom-style="CHANGE">followed by lower levels measured after</span> the last training session of the 12 wk interventions (31 sessions)
[@RN2358].
Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT
[@RN2358].
These data suggest that ribosome accumulation reaches a plateau in the early phase of RT and that increases are sensitive to training volume in constant volume protocols.

@@

For both ribosomal protein S6 (RPS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with RPS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and RPS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For RPS6, de-training did not affect protein levels, which remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while RPS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

@@

In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1). In contrast, no evidence was found for a relationship between total RNA and RPS6 protein levels (Table 1).

@@

cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")
```
<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for RPS6 and UBF, respectively.</span>

>>>>

<div custom-style="runningtitle_style">**Short title:** Ribosome biogenesis in early phase resistance training</div>

@@

**Aim:** To describe <span custom-style="CHANGE">ribosome biogenesis during resistance training, its relation to training volume and muscle growth</span>.

**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. <span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline, 48 hrs after sessions 1, 4, 5, 8, 9, and 12,</span> and after eight days of de-training, and from a control group (n = 8). Muscle thickness was measured before and after the intervention</span>.

**Results:** Training led to muscle growth (`r hypertrophy` vs. control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions.<span custom-style="CHANGE"> Total RNA increased rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline). Total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy</span> (`r rna_hypertrophy` per %-point increase in total RNA per session). After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.

@@

**Keywords:** Muscle hypertrophy, Resistance training, Ribosome biogenesis

@@

Skeletal muscle is a critical target for interventions that promote health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
with resistance training (RT) being the advocated remedy.  
Prolonged RT leads to changes in the balance between muscle protein breakdown and synthesis, with one bout of resistance exercise acutely increasing protein synthesis for up to 48 hrs after exercise [@RN786],<!-- Phillips 1997 -->
<span custom-style="CHANGE">and</span> subsequent repeated bouts leading to accumulation of muscle protein over time
[@RN1272;@RN1642]<!-- Phillips review 2014 --><!-- Brook 2015 experiment -->. 
In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased <span custom-style="CHANGE">basal muscle protein synthesis rates</span>
[@RN1521;@RN1866;@RN1897],<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --><!-- Wilkingson 2008 --><!-- Reidy 2017 Increased rested state protein synthesis -->
which has been postulated to be associated with increased translational capacity, i.e. accumulation of ribosomes
[@RN1897;@RN1912]<!-- Reidy 2017 --><!-- Figueiredo 2017 Ribosome biogenesis review -->.
This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, which is closely connected to protein synthesis
[@RN1754;@RN2145]<!-- West 2016 correlation between total RNA and protein synthesis in C2C12--><!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
and muscle hypertrophy
[@RN1644;@RN1755;@RN2358].<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp --><!-- Stec Ribosome bio higher in high responders --><!-- Volume dependent RNA accum relates to volume depend hyper -->
Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
[@RN1755;@RN1754;@RN1810]<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty --><!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS--><!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.

Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
[@RN1920;@RN1820;@RN1810]<!-- Warner review yeast --><!-- Moss review --><!-- Inhibition of Pol II affects trasncription at Pol I -->.
Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
[@RN1820]<!-- Moss review -->.
Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
[@RN2582;@RN2563].<!-- Tuan UBF phosphorylation important for SL1-UBF interaction --><!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
Such activation is at least partly controlled by the mechanosensitive mTOR pathway, with its inhibition being associated with blocked UBF phosphorylation and subsequent rRNA transcription
[@RN1632;@RN2564].<!-- Nader rapamycin block UBF availability  --><!-- Hannan mTOR  -->
Interestingly the availability of UBF *per se* has been shown to be a determinant of rRNA transcription
[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->

Resistance exercise is a potent <span custom-style="CHANGE">and specific [@figueiredo2021]</span> stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
[@RN1656;@RN1037].<!-- A single session upregulates 45 S 24 h post ex --><!-- Nader 2014 45S up in response to 4 h after acute exercise -->
Repeated bouts lead to the accumulation of mature rRNA reflected in total RNA and presumably functional ribosomes
[@RN1656;@RN1520;@RN2358;@RN1755;@RN1644;@RN1809;@RN1897].<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old --><!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr--><!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young--><!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD--><!-- Figuieredo 2015 16 sessions 49% increase young --><!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young--><!-- 11% 36 sessions young -->
However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with a mere few studies having investigated exercise-induced changes in rRNA over multiple time-points, all of which are either limited to a selected few time-points or a limited time frame.
For example, two consecutive bouts of electrically evoked muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
[@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
Using voluntary contractions, peak values were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809],
resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), <span custom-style="CHANGE">followed by lower levels measured after</span> the last training session of the 12 wk interventions (31 sessions)
[@RN2358].
Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT
[@RN2358].
These data suggest that ribosome accumulation reaches a plateau in the early phase of RT and that increases are sensitive to training volume in constant volume protocols.

@@

For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12, and remained elevated after eight days of rest (Figure 2A and B), with no robust differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For rpS6, de-training did not affect protein levels, which remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

@@

In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1). In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).

@@

cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")
```
<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for rpS6 and UBF, respectively.</span>
